### 2014-03-17 thesis-proposal/problem.tex
Updated ICSE14 and proposal

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
%\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
%and generality in language design is a serious unsolved 
%problem. }
%- Reynolds, 1970\todo{citations}\end{quote}
%%One might exp ``generality''
Designing and implementing a simple but general-purpose programming language and its associated tools (collectively, a \emph{programming system}) has long been seen as a grand challenge in computing. Were an ideal system to emerge, we might expect that its experienced users would not often see the need to construct a new dialect of the language or its tools to implement a new high-level abstraction. A library-based implementations of an abstraction would be essentially as reasonable, natural and performant as a standalone implementation. By this metric, we are clearly far from our ideal. New language and tool dialects, originating both in research and in practice and spanning all major language paradigms, continue to be widely used as vehicles for new abstractions (examples of which we will discuss throughout this work), suggesting that not all useful  abstractions can be seen, when considered comprehensively, as mere modes of use of a contemporary ``general-purpose'' programming system.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.

We observe that new system dialects are often motivated by the desire to introduce new syntax, strengthen the language's type system or provide specialized editor services, all aspects of existing systems that appear rather monolithic when considered from the perspective of a library provider. %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally. 
It follows, then, that {language-integrated mechanisms} that allow library providers to extend these aspects of the system would decrease the need for new language and tool dialects, increasing the generality of the  system. However, such mechanisms must be introduced with care, because decentralizing control over these fundamental aspects of the system could destabilize it by weakening its metatheory, permitting ambiguities and conflicts and obscuring the semantics of a piece of code. 
In this thesis, we aim to show that by associating language extensions with user-defined types (forming what we call \emph{active types}) and structuring the underlying extension mechanism to enforce critical  abstraction barriers, we can give library providers the ability to express powerful extensions to the syntax, the type system and even to editor services from within libraries, while avoiding these kinds of stability issues. %We call user-defined types that introduce new features into the system in this way \emph{active types}.  % also highly expressive.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
\subsection{Motivating Example: Regular Expressions}\label{regex}
To make the problems we aim to address more concrete, we begin with a simple example that we will return to throughout this work. \emph{Regular expressions} are commonly used to express patterns in strings (e.g. DNA sequences) \cite{Thompson:1968:PTR:363347.363387}. If a programming system provided full support for this abstraction, it might  provide, or enable the provision of, features like:

\begin{enumerate}
\item \textbf{Syntax for pattern literals}. An ideal syntax would permit us to express patterns concisely and clearly. For example, The BisI restriction enzyme cuts DNA whenever it sees the pattern \texttt{GC\textit{N}GC}, where \texttt{\textit{N}} represents any base. We would want to express it, perhaps, as follows (using curly braces to splice one pattern into another):
\begin{lstlisting}[numbers=none]
let N : Pattern = <SURLA|T|G|CEURL>
let BisI : Pattern = <SURLGC{EURLNSURL}GCEURL>\end{lstlisting}
%The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
Malformed patterns would result in intelligible {compile-time} errors.
\item A \textbf{type system} that ensures that key invariants related to regular expressions are statically maintained:
	\begin{enumerate}
	\item only other patterns and properly escaped strings are spliced into a pattern, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
	\item out-of-bounds backreferences to a captured group are not used \cite{spishak2012type}
	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured as membership in a regular language \cite{fulton-thesis}
	\end{enumerate}
%When an error is found, an intelligible error message is provided.
%\item An \textbf{implementation} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time. In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs. If the developer is not careful to cache compiled representations, regular expressions used repeatedly in a program might be needlessly re-compiled on each use. %By performing this step ahead-of-time, these dangers can be avoided.
\item \textbf{Editor services} that allow clients to refer to documentation, interactively test patterns against test strings and search for common patterns (e.g. \cite{IntelliJRegexp})
\end{enumerate}

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today builds in support for all of the features enumerated above in their strongest form, so library providers must provide support for regular expressions by leveraging  general-purpose constructs. Unfortunately, it is impossible to completely define the syntax and the specialized static semantics described in the references above in terms of general-purpose notations and abstractions. Library providers thus need to compromise.

@@

None of these approaches address the issue of tool support (feature 3). Regular expressions are not trivial to work with, and we have found that tool support can be quite helpful, even for professional programmers \cite{Omar:2012:ACC:2337223.2337324}. A number of tools are available online that provide helpful services like syntax highlighting for patterns and pattern extraction from examples (e.g. \cite{_txt2re:_????}). Unfortunately, tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}. They are also less usable than editor-integrated tools (e.g. \cite{IntelliJRegexp}) because they require the programmer to switch contexts  and cannot make use of the editor context. Working with high-level abstractions in a simple text editor is thus awkward and error-prone, just like attempting to work with high-level abstractions in a strictly low-level language. For these reasons, we take the position that the editor is as integral to a good programming system as the language itself (though not the position that it be a part of a full integrated development environment).
\subsection{Language-External Approaches}\label{external-approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=.48]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with orthogonality and client compatibility (described in the text). (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active'' libraries. If the extension mechanism guarantees that extensions cannot cause instabilities, the problems of orthogonality and client compatibility are avoided.}
\label{approaches}
%\vspace{-10px}
\end{figure}

@@

%This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
%But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. 
As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for a type introduced by F\#, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET. This is not an issue exclusive to intermediate languages that make regrettable choices regarding \verb|null|, however. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this more specialized static invariant is left entirely unchecked at language boundaries. Exposing functions that operate over datatypes, tuples and values having units of measure is not recommended when a component ``might be used'' from another language, according to drafted F\# component design guidelines \cite{fsharpguidelines}, because it is awkward to construct and consume these from other languages without the convenient primitive operations (e.g. pattern matching) and syntax that F\# includes. SML.NET prohibits exposing such types at component boundaries altogether (and cannot naturally consume F\# data structures, despite having a rather similar syntax and semantics in many ways). 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
%In some cases, desirable features must be omitted entirely due to concerns about interoperability. F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no apparent analogs in the type system of the CLI.
%\end{itemize}

>>>>

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
%\begin{quote}\textit{The recent development of programming languages suggests that the simultaneous achievement of simplicity 
%and generality in language design is a serious unsolved 
%problem. }
%- Reynolds, 1970\todo{citations}\end{quote}
%%One might exp ``generality''
Specifying and implementing a simple but general-purpose programming language and its associated tools (collectively, a \emph{programming system}) has long been seen as a grand challenge in computing. Were an ideal system to emerge, we might expect that providing a new high-level abstraction would not require specifying a new dialect of the language or its tools, because a library-based embedding of the abstraction would be essentially as reasonable, natural and performant. By this metric, we are clearly far from our ideal. New language and tool dialects, originating both in research and in practice and spanning all major language lineages, continue to be widely used as vehicles for new abstractions (examples of which we will discuss throughout this work). This suggests that not all useful  abstractions can be seen, when considered comprehensively, as mere modes of use of a contemporary ``general-purpose'' programming system.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved.

We observe that new dialects are often motivated by the desire to introduce new syntax, strengthen the language's type system or provide specialized editor services, all aspects of existing systems that appear rather monolithic when considered from the perspective of a library provider. %, encouraged  historically  by the availability of tools like compiler generators and,  more recently, language workbenches \cite{workbenches} and DSL frameworks \cite{dsl}. Unfortunately, taking this approach makes it substantially more difficult for clients to import high-level abstractions orthogonally. 
It follows, then, that {language-integrated mechanisms} that allow library providers to extend these features of the system would increase its generality by decreasing the need for new language and tool dialects. However, such mechanisms must be introduced with care, because decentralizing control over such fundamental aspects of the system could destabilize it by weakening its metatheory and permitting ambiguities and conflicts (collectively, \emph{safety issues}). It could also make it more difficult to reason about the meaning of code. 
In this thesis, we aim to show that by adopting the natural convention of organizing language extensions around user-defined types (forming what we call \emph{active types}) and designing an underlying type system that enforces critical  abstraction barriers between extensions, we can give abstraction providers the ability to express from within libraries powerful, orthogonal extensions to the syntax and type system, and define powerful editor services, without the possibility of the kinds of safety issues that have limited prior approaches. %We call user-defined types that introduce new features into the system in this way \emph{active types}.  % also highly expressive.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
\subsection{Motivating Example: Regular Expressions}\label{regex}
To make the problems we aim to address more concrete, we begin with a simple example that we will return to throughout this work. \emph{Regular expressions} are commonly used to express patterns in strings (e.g. DNA sequences) \cite{Thompson:1968:PTR:363347.363387}. If a programming system provided full support for this abstraction, it might  provide features like:

\begin{enumerate}
\item \textbf{Syntax for pattern literals}. An ideal syntax would permit us to express patterns concisely and clearly. For example, The BisI restriction enzyme cuts DNA whenever it sees the pattern \texttt{GC\textit{N}GC}, where \texttt{\textit{N}} represents any base. We would want to express it, perhaps, as follows (using curly braces to splice one pattern into another):
\begin{lstlisting}[numbers=none]
let N : Pattern = <SURLA|T|G|CEURL>
let BisI : Pattern = <SURLGC{EURLNSURL}GCEURL>\end{lstlisting}
%The cognitive load of reading and writing patterns is low and patterns are parsed once at compile-time. 
Malformed patterns would result in intelligible {compile-time} errors.
\item A \textbf{type system} that ensures that key invariants related to regular expressions are statically maintained:
	\begin{enumerate}
	\item only other patterns and properly escaped strings are spliced into a pattern, to avoid splicing errors and injection attacks \cite{owasp2013, Bravenboer:2007:PIA:1289971.1289975}
	\item out-of-bounds backreferences to a captured group are not used \cite{spishak2012type}
	\item string processing operations do not lead to a string that is malformed, when well-formedness can be captured as membership in a regular language \cite{fulton-thesis}
	\end{enumerate}
%When an error is found, an intelligible error message is provided.
%\item An \textbf{implementation} that partially or fully compiles known regular expressions into the efficient internal representation that will be used by the regular expression matching engine (e.g., a finite automata \cite{Thompson:1968:PTR:363347.363387}) ahead of time. In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing performance overhead into programs. If the developer is not careful to cache compiled representations, regular expressions used repeatedly in a program might be needlessly re-compiled on each use. %By performing this step ahead-of-time, these dangers can be avoided.
\item \textbf{Editor services} to support syntax highlighting of patterns and to allow clients to refer to documentation and interactively test patterns against examples strings (e.g. \cite{IntelliJRegexp}) or even extract patterns from examples interactively (e.g. as in \cite{_txt2re:_????}).
\end{enumerate}

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today builds in support for all of the features enumerated above in their strongest form, so library providers must provide support for regular expressions by leveraging  general-purpose constructs. Unfortunately, it is impossible to completely define the syntax and the specialized static semantics described in the references above in terms of general-purpose notations and abstractions, so library providers need to compromise.

@@

None of these approaches address the issue of tool support (feature 3). Regular expressions are not trivial to work with, and we have found that tool support can be quite helpful, even for professional programmers \cite{Omar:2012:ACC:2337223.2337324}. A number of tools are available online that provide helpful services like syntax highlighting for patterns and pattern extraction from examples (e.g. \cite{_txt2re:_????}). Unfortunately, tools that must be accessed externally are difficult to discover and are thus used infrequently \cite{Murphy-Hill:2011:PIE:1958824.1958888,Campbell:2008:DRT:1636642.1636651,Omar:2012:ACC:2337223.2337324}. We found in experiments that they are also less usable than editor-integrated tools %(e.g. \cite{IntelliJRegexp}) 
because they require the programmer to switch contexts  and cannot make use of the code context \cite{Omar:2012:ACC:2337223.2337324}. Working with high-level abstractions in a simple editor is thus awkward and error-prone, just like attempting to work with high-level abstractions in a low-level language. For these reasons,   we consider editor behavior as within the purview of an abstraction specification.
\subsection{Language-External Approaches}\label{external-approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=.48]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) When taking a language-external approach, new features are packaged together into separate languages and tools, causing problems with orthogonality and client compatibility (described in the text). (b) When taking a language-integrated approach, there is one extensible host language and the compile-time and edit-time logic governing new constructs is expressed within ``active'' libraries. If the extension mechanism guarantees that extensions cannot cause instabilities, the problems of orthogonality and client compatibility are avoided.}
\label{approaches}
%\vspace{-10px}
\end{figure}

@@

%This approach only works well when new languages consist of constructs that can also be expressed safely and almost as naturally in the common language.
%But many of the most innovative constructs found in modern languages (often, those that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. 
As a simple example with significant contemporary implications, F\#'s type system does not admit \verb|null| as a value for a type introduced by F\#, but maintaining this sensible internal invariant still requires dynamic  checks because the stricter typing rules of F\# do not apply when F\# data structures are constructed by other languages on the Common Language Infrastructure (CLI) like C\# or SML.NET. This is not an issue exclusive to intermediate languages that make regrettable choices regarding \verb|null|, however. The F\# type system also includes support for checking that units of measure are used correctly \cite{syme2012expert, kennedy1994dimension}, but this more specialized static invariant is left entirely unchecked at language boundaries. Exposing functions that operate over datatypes, tuples and values having units of measure is not recommended when a component ``might be used'' from another language \cite{syme2012expert} because it is awkward to construct and consume these from other languages without the convenient primitive operations (e.g. pattern matching) and syntax that F\# includes. SML.NET prohibits exposing such types at component boundaries altogether. It cannot it naturally consume F\# data structures, despite having a rather similar syntax and semantics in many ways (both languages directly descend from ML). 
%In Scala, traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. 
%In some cases, desirable features must be omitted entirely due to concerns about interoperability. F\#, for example, aimed to retain source compatibility with Ocaml code, but due to the need for bidirectional interoperability with CLI languages, it does not support features like polymorphic variants, modules or functors \cite{ocaml-manual} because they have no apparent analogs in the type system of the CLI.
%\end{itemize}

### 2014-03-17 thesis-proposal/timeline.tex
Updated ICSE14 and proposal

% !TEX root = omar-thesis-proposal.tex
\section{Timeline}
\begin{itemize}
\item @$\lambda$ - ICFP - Mar. 1
\item Wyvern - ECOOP or OOPSLA - late March
\item Ace - PLDI or OOPSLA - late March
\item Then write everything up.
\end{itemize}

>>>>

% !TEX root = omar-thesis-proposal.tex
\section{Conclusion}\label{conclusion}
In arguing for language extensibility, we are accepting Carnap's logical pluralism\todo{citations}. But we take issue with his \emph{principle of tolerance}, which he explained as follows:
\begin{quote}
In logic, there are no morals. Everyone is at liberty to build his own logic, i.e. his own form of language, as he wishes. All that is required of him is that, if he wishes to discuss it, he must state his methods clearly, and give syntactic rules instead of philosophical arguments.
\end{quote}
%Aside from the archaic use of gendered pronouns, 
Carnap formulated his principle of tolerance because he wished to freely explore the consequences of reasoning with different connectives and  rules and emphasize that conducting philosophical reasoning by methodical logical analysis did not require one to accept any particular axioms. But he was also a prominent member of the Vienna circle, which sought a ``unified science'' the purpose of which was "to link and harmonise the achievements of individual investigators in their various fields of science" by the construction of a "constitutive system" \cite{ViennaCircle}. These two programs were in conflict. If the reader will permit a brief digression into social science, the adoption of a tolerant viewpoint is difficult to sustain in a society that does not require \emph{reciprocity}. Yossie Nehushtan goes as far as to argue that a liberal  society ``should not tolerate anything that denies the justifications of tolerance'' \cite{Nehushtan}. According to Raz, the primary justification of tolerance is to maximize 
individual autonomy in a multi-party context \cite{Raz}. And indeed, this was precisely Carnap's program. But were the Vienna circle's ``constitutive system'' to become na\"ively tolerant of new inference rules (what Carnap called L-rules), it would find that this would then deny its investigators autonomy of reasoning. Thus, we propose a more cooperative \emph{principle of reciprocal tolerance}:
\begin{quote}
In metalogic, there \emph{are} morals. Everyone is at liberty to build their own logical fragment, i.e. their own language extensions, as long as these can be clearly shown, by formal means instead of philosophical argument, to conserve autonomously derived reasoning principles when used in \emph{any} combination.
\end{quote}

By following this principle, we believe that a research program that supports Carnap's logical pluralism within a ``constitutive system'' of the sort envisioned by Vienna circle is both possible and practical. The work in this thesis gives first steps towards this goal.

% http://www.trinitinture.com/documents/nehushtan.pdf
% https://en.wikipedia.org/wiki/Vienna_Circle
% http://consequently.org/papers/carnap.pdf
% https://www.academia.edu/4041290/How_tolerant_can_you_be_Carnap_on_the_normativity_of_logic

\section{Timeline}
To summarize, we plan on completing the work in this thesis per the following schedule:
\begin{itemize}
\item The work on type-specific languages has been accepted to ECOOP 2014. The deadline for final revisions is May 12, so we will complete the remaining work by then.
\item The work on Ace will be submitted to OOPSLA 2014 on Mar. 25.
\item The work on @$\lambda$ and active embeddings is substantially complete. We plan to complete some technical details and exposition after Mar. 25 and submit it to POPL in July (though we hope to be done with the thesis work before then).
\item The work on active code completion has been completed and published at ICSE 2012 \cite{Omar:2012:ACC:2337223.2337324}. The remaining work is expository and will be completed in the course of writing the dissertation.
\end{itemize}
